Cross-Loop Optimization of Arithmetic Intensity for Finite Element Local Assembly

### 2014-09-09 TACO-paper-v2/coffee.tex
Section 6: copy editing

We observe that our choice was to develop COFFEE as a separate, self-contained software module, with clear input/output interfaces, rather then incorporating it within PyOP2. This was motivated by two critical aspects that characterize the generality of our research.

\paragraph{Separation of concerns} We believe that in domain-specific frameworks there must be a clear, logical separation of roles reflecting the various levels of abstractions, with domain-specialists that can completely abstract from performance optimization. In Firedrake, for instance, COFFEE decouples the mathematical specification of a finite element method, captured by the Unified Form Language and the FEniCS Form Compiler, from code optimization. This is of fundamental importance to maximize productivity by allowing scientists to focus only on their area of expertise. Practically speaking, from the perspective of the domain-specific language and compiler designers, our optimization strategy represents an incentive to produce extremely simple representations of the code (e.g. fully-inlined mathematical expressions in the form of an abstract syntax tree, in the case of Firedrake) so as to make the architecture-aware code optimizer completely responsible for choosing and applying the most suitable set of transformations.

\paragraph{Generalizability to other domains} There are neither conceptual
nor technical reasons which prevent our compiler being used in applications
other than Firedrake. For example, integration with the popular FEniCS
framework, the pioneer of automated code generation for finite element local
assembly, would be relatively easy to achieve. It is more challenging to
assess the generality of the optimization strategy: the extent to which COFFEE and its transformations are transferable to other computational domains, perhaps other DSLs, and to what extent this would be helpful for improving full-application performance. To answer these questions, we first need to go back to the origins of our compiler. The starting point of our work was the mathematical formulation of a local assembly operation, expressible as follows

@@

\item In Equation~\ref{eq:assembly-model}, the summations correspond to reduction loops, whereas loops over indices $i$ and $j$ are fully parallel. Throughout the paper we assumed a kernel be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs. On the other hand, we note that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability.

@@

  successfully applied to improve register pressure. However, underlying

@@

  mesh applications has been adopted in~\cite{carlolcpc}). In
  Equation~\ref{eq:assembly-model}, for instance, not only can  any of the

@@

  alignment are special cases, since they could be performed at run-time if some values were not known at code generation time. Theoretically, they could also be automated by a general-purpose compiler though profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.

>>>>

COFFEE was developed as a separate, self-contained software module, with
clear input/output interfaces, rather then incorporating it within
PyOP2. This choice was motivated by two critical aspects that
characterize the generality of our research.

\paragraph{Separation of concerns} We believe that in domain-specific
frameworks there must be a clear, logical separation of roles reflecting
the various levels of abstraction, where domain specialists are
completely separated from performance optimization. In Firedrake, for
instance, COFFEE decouples the mathematical specification of a finite
element method, captured by the Unified Form Language and the FEniCS
Form Compiler, from code optimization. This is of fundamental importance
to maximize productivity by allowing scientists to focus only on their
area of expertise. Practically speaking, from the perspective of the
domain-specific language and compiler designers, our optimization
strategy represents an incentive to produce extremely simple
representations of the code (e.g. fully-inlined mathematical expressions
in the form of an abstract syntax tree, in the case of Firedrake) so as
to make the architecture-aware code optimizer completely responsible for
choosing and applying the most suitable set of transformations.

\paragraph{Generalizability to other domains} There are neither
conceptual nor technical reasons which prevent our compiler from being
used in applications other than Firedrake. For example, integration with
the popular FEniCS framework, the pioneer of automated code generation
for finite element local assembly, would be relatively easy to achieve.
It is more challenging to assess the generality of the optimization
strategy: the extent to which COFFEE and its transformations are
transferable to other computational domains, perhaps other DSLs, and to
what extent this would be helpful for improving full-application
performance. To answer these questions, we first need to go back to the
origins of our compiler. The starting point of our work was the
mathematical formulation of a local assembly operation, expressible as
follows

@@

\item In Equation~\eqref{eq:assembly-model}, the summations correspond
  to reduction loops, whereas loops over indices $i$ and $j$ are fully
  parallel. Throughout the paper we assumed a kernel to be executed by a single thread, which is likely to be the best strategy for standard multi-core CPUs. On the other hand, we note that for certain architectures (e.g. GPUs) this could be prohibitive due to memory requirements. Intra-kernel parallelization is one possible solution: a domain-specific compiler such as COFFEE could map mathematical quantifiers and operators to different parallelization schemes and generate distinct variants of multi-threaded kernel code. Based on our experience, we believe this is the right approach to achieve performance portability.

@@

  successfully applied to improve register pressure. However, the underlying

@@

  mesh applications has been adopted in~\citeN{carlolcpc}). In
  Equation~\eqref{eq:assembly-model}, for instance, not only can  any of the

@@

  alignment are special cases, since they could be performed at run-time
  if some values were not known at code generation time. Theoretically,
  they could also be automated by a general-purpose compiler through profile-guided optimization, provided that some sort of data-flow analysis is performed to ensure that the extra loop iterations over the padded region do not affect the numerical results.
